---
title: "Exercise I, PostDay1"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
Weakliem<-read.table('eppsac-utdallas/data/Weakliem.txt', header=TRUE)

library (MASS)
library(mgcv)
attach(Weakliem)
```

## The data

Each case represents a country. The variables are:

– **secpay**: Average public opinion in the country towards
pay inequality. High values indicate egalitarian attitudes.
(From the World Values Survey)
– **gini**: The gini coefficient (a measure of income
inequality).
– **gdp**: The gross domestic product for the country.
– **hetero**: A measure of the level of ethnic and religious
heterogeneity. High values indicate highly homogeneous.
– **democrat**: Coded 1 if the country is democratic; 0 if it is
not



1. Using the `gam` function, fit a generalized additive model
regressing `secpay` on `gini` and `gdp` (both with a
smoothing term). Using the `summary` function, see if the
smooth terms are statistically significant?

2. Use the `plot` function to plot the trends and their
standard errors. Try plotting the on the same page.
Type ?plot.gam for help.

3. Continuing from the previous exercise, fit a semiparametric
model regressing `secpay` on `gini` and
`democrat`. (Remember, `democrat` is a categorical
variable). Explore the summary for the model to see if
democracy is statistically significant.

4. Now fit a model that specifies an interaction between
democrat and s(gini). (You will need to create two
dummy variables for dummy—Type ?gam for help).

5. Plot the effects of the above interaction, putting the two
smooths on the same page (type ?split.screen and
?plot.gam for help).

6. Detach the `mgcv` package (but not the Weakliem
dataset), and load the `gam` package. [We detach mgcv
when before loading the gam package because they have
different functions of the same name]. Type ?detach for
help.

7. Now refit the model from exercise 4 above, specifying
a lowess smooth for the gini effect (type ?lo for help).
Compare the results of this model to those in exercise
4.

```{r}
mod.gam1<-gam(secpay~s(gini)+s(gdp))
summary(mod.gam1)

#4
plot(mod.gam1, se=TRUE, pages=1)

#5
mod.semi1<-gam(secpay~s(gini)+democrat)
summary(mod.semi1)

#6
  ##must create two separate dummy variables for the
  ##categorical variable to be used in an interaction
demo<-as.numeric(democrat==1)
nondemo<-as.numeric(democrat==0)
  ##gam including democracy*s(gini) interaction
mod.semi2<-gam(secpay~s(gini, by=demo)+s(gini, by=nondemo))
summary(mod.semi2)

#7
split.screen(figs=c(1,2))
screen(1)
plot(mod.semi2, select=1)
title("Democracies")
screen(2)
plot.gam(mod.semi2, select=2)
title("Nondemocracies")
close.screen(all=TRUE)

#8
detach(package:mgcv)
library(gam)

#9
mod.lo<-gam(secpay~lo(gini)+democrat)
summary(mod.lo)
plot(mod.lo, se=TRUE) 
```


#Exploratory Data Analysis

```{r, eval =FALSE}
#Ex. 1
# Text files are read using the "read.table" function
# header=TRUE treats the first row as variable names

attach(Weakliem)
summary(Weakliem)

#Ex. 2
# loading the "car" library
library(car)
scatterplotMatrix(~secpay + gini + gdp+ hetero | democrat)

#Ex. 3
plot(gini, secpay)
abline(lm(secpay~gini)) #plots line from linear model

plot(gdp, secpay)
abline(lm(secpay~gdp))


#Ex. 4
split.screen(figs=c(1,2))
screen(1)
plot(gini,secpay)
abline(lm(secpay~gini))
screen(2)
plot(gdp, secpay)
abline(lm(secpay~gdp))
close.screen(all=TRUE)

    ##Alternative way using "par" and "mfrow"
    par(mfrow=c(1,2))  #specifies 1 row and 2 columns
    plot(gini,secpay)
    abline(lm(secpay~gini))
    plot(gdp, secpay)
    abline(lm(secpay~gdp))
    
##Ex. 5
mod1<-lm(secpay~gdp+gini)
summary(mod1)
par(mfrow=c(2,2)) #puts all 4 diagnostic plots on one screen
plot(mod1)

##Ex. 6
par(mfrow=c(1,1)) #setting the screen back to one display
crPlots(mod1) # component-plus-residuals plots in car package
     #to exit, type 0

##Ex. 7
mod2<-lm(secpay~gini+democrat)
summary(mod2)

##Ex. 8
democrat2<-as.factor(democrat) #as.factor treats democrat as categorical
mod3<-lm(secpay~gini*democrat2)  
summary(mod3)

##Ex. 9
if (require(effects))
  plot(all.effects(mod3))

##Ex. 10
detach(Weakliem)

############################################
###Scatterplots and Transformation Exercises
############################################

##1
library(car)
data(Prestige) #loads a dataset that is included in a package
attach(Prestige)

##2
scatterplot(income, prestige)

##3
log.inc<-log(income)
scatterplot(log.inc, prestige)

##4
scatterplot(prestige~income|type)


##############################
###Polynomial Regession
##############################

##1 and 2
mod2<-lm(prestige~poly(income,3)+type)
summary(mod2) ##quadratic model fits best:
mod3<-lm(prestige~poly(income,2)+type)

#3
income2<-income^2 
mod3b<-lm(prestige~income+income2+type)
summary(mod3)
summary(mod3b)

#4
mod4<-lm(prestige~income^2+type)

AIC(mod3)
AIC(mod4)
    ##Mod 3 fits much better
if (require(effects)){
  plot(all.effects(mod3))
  plot(all.effects(mod4))
}



###############################
##Lowess and Smoothing Splines
###############################

#1
library(car)
data(Prestige)
attach(Prestige)
mod.lm<-lm(prestige~income)
mod.lowess<-lowess(income, prestige)
mod.spline<-smooth.spline(income, prestige)

#2
plot(income,prestige)
abline(mod.lm, col="blue")
lines(mod.lowess)
lines(mod.spline,col="red")

#3
mod.lm<-lm(prestige~income)
mod.loess<-loess(prestige~income)
summary(mod.loess)

    #residual sum of squares for the two models
rss1<-sum(residuals(mod.lm)^2)
rss2<-sum(residuals(mod.loess)^2)
    #F-Test
Ftest<-((rss1-rss2)/(6.09-2))/(rss2/(102-6.09))
pf(Ftest, 6.09-2, 102-6.09, lower.tail=FALSE)
    ##significant difference


```

