---
title: "San Francisco Crime Data"
author: "M Loecher"
output: 
  html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(lubridate)
library(dygraphs)
library(xts)
library(mgcv)
revealCode=FALSE
```


Read in sample of SF crime data for 2014

```{r, echo=TRUE}
print(load("../data/crimes2014.rda"))#

class(crimes2014$Time)

head(crimes2014)

```

### date/time classes in R


1. How would you compute the *DayOfWeek* from the timestamp? (Hint: check out the library *lubridate*)
 
```{r, echo=revealCode}
library(lubridate)
crimes2014$DayOfWeek = as.character(wday(crimes2014$Time,label=TRUE))
crimes2014$DayOfWeek = wday(crimes2014$Time,label=TRUE)
```
 
2. How would you compute the *HourOfWeek* from the timestamp?
What ambiguity needs to be first agreed upon so that you can compare your results to the data given.
```{r, echo=revealCode}
crimes2014$TimeOfDay = hour(crimes2014$Time) + minute(crimes2014$Time)/60
crimes2014$HourOfWeek = crimes2014$TimeOfDay + (wday(crimes2014$Time)-1)*24

```

2. Compute and graph the integer weekhour pattern of violent crime rate as a barplot. The function *table* will be useful. An example is shown below.

```{r, echo=revealCode, fig.width=12}
DoW=c("Sunday","Monday","Tuesday","Wednesday","Thursday","Friday","Saturday")

  m = table(round(crimes2014$HourOfWeek),crimes2014$violent)
  bp =  barplot(m[,"TRUE"]/rowSums(m))
  
  bp =  barplot(m[,"TRUE"]/rowSums(m), xlab = "hour of week", ylab="violent proportion", col = "blue",space = 0,names.arg="")
  abline(v= bp[seq(24,by=24,length=6)], lty = 2, col="green")
  
  for (j in 1:7) mtext(substring(DoW, 1,3)[j], side=3, at =bp[seq(12,by=24,length=7)[j]], col = "green")
```



3. Using the commands *table()* and *format()* build an hourly time series of counts of (all) crimes. (Hint: as.POSIXct("2012-01-02 08", format="%Y-%m-%d %H"))
```{r, echo=revealCode}
h = table(format(crimes2014$Time,  "%Y-%m-%d %H:00:00"))
```

4. Plot that time series.

```{r,echo=revealCode}
plot(as.POSIXct(names(h)), as.numeric(h), type="l", xlab = "", ylab = "Num Crimes", col = "darkblue")
title("Hourly crimes in San Francisco")
grid()
```


### Interactive Charts

5. Use the package *dygraphs* and look up the documentation
at http://rstudio.github.io/dygraphs/
Try to understand the pipe operator %>% and create another interactive time series chart with a range selector!

```{r echo=revealCode}
 
#CrimeCounts=xts(NumCrimes=as.numeric(h),as.POSIXct(names(h))) 
CrimeCounts=xts(as.numeric(h),as.POSIXct(names(h)))
plot(CrimeCounts)
#dygraph(CrimeCounts)

dygraph(CrimeCounts) %>%  dyRangeSelector() 
```

6. The * %>% *command is a so called "pipe operator" which is super useful in that it passes the   output from one command to another. Simply left to right! In base R we would normally achieve  this by chaining functions, e.g. in class we often nest commands such as 
 ``` round(mean(x),2) ```,  which in pipe notation would then be written as ``` mean(x) %>% round(,2) ```



# GAMs 

Generalized Additive Models to model temporal patterns.

7. Fit a GAM to find the hour-of-day pattern

```{r,echo=revealCode}

#better illustrated with crime data!
fit5 = gam(violent ~ s(TimeOfDay, k = 12), family=binomial, data = crimes2014)
plot(fit5, trans=exp)
summary(fit5)
fittedVals = predict(fit5, type = "response", se.fit = TRUE)

```

8. Fit a GAM to find the day-of-week pattern

9. Fit a GAM that combines both effects additively

10. 
   * Do you believe that this would be a sufficiently rich model ? 
   * How else could you combine the two different periodicities ?
   * Look up cyclic spines (periodic boundary conditions) and decide if their implementation makes sense for this case.
   * Look up the help function for `bam` and try to speed up your fit.

## Temporal Correlations

11. model the autocorrelation!


