---
title: "Nonlinear Modeling II"
author: "M Loecher"
output:
   pdf_document:
     extra_dependencies: ["pdfpages"]
     toc: true
     toc_depth: 2
     includes:
        in_header: header.tex
classoption: landscape
header-includes: \usepackage{amsmath}
subtitle: Generalized Additive Models (close to chapter 7, ISLR)
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)

#library(gamair)
data(mpg, package="gamair")
library(lubridate)
library(knitr)
#library(dygraphs)
library(xts)
library(ISLR)
data(Auto)
data(mcycle,package="MASS")
library(splines)
library(boot)
library(mgcv)
```

\includepdf[pages={2}]{../datacamp/chapter1.pdf}

# Pros and Cons of GAMs

\benum

\item[+] GAMs allow us to fit a non-linear $f_j$ to each $X_j$, so that we can
automatically model non-linear relationships that standard linear regression
will miss. This means that we do not need to manually try
out many different transformations on each variable individually.
\item[+] The non-linear fits can potentially make more accurate predictions
for the response Y .
\item[+] Because the model is additive, we can still examine the effect of
each $X_j$ on Y individually while holding all of the other variables
fixed. Hence if we are interested in inference, GAMs provide a useful
representation.

\item[+] The smoothness of the function $f_j$ for the variable $X_j$ can be summarized
via degrees of freedom.

\item[-] The main limitation of GAMs is that the model is restricted to be
additive. With many variables, important interactions can be missed.
However, as with linear regression, we can manually add interaction
terms to the GAM model by including additional predictors of the
form $X_j × X_k$. In addition we can add low-dimensional interaction
functions of the form $f_{jk}(X_j,X_k)$ into the model; such terms can
be fit using two-dimensional smoothers such as local regression, or
two-dimensional splines (not covered here).

\eenum

\includepdf[pages={3-7,10-15}]{../datacamp/chapter1.pdf}

## [GAM Exercises part I](exercises/GAM-Exercises-I.html)

# Multiple Smooths

A true understanding of the label *additive* only comes in the context of multiple variables. A GAM is an extension of the multiple linear regression model which allows for non-linear relationships between each feature and the response. We simply replace each linear component $\beta_j x_{ij}$ with a (smooth) nonlinear function $f_j( x_{ij} )$ so that the model becomes:

$$
\label{eq:gamDef}
y_i = \beta_0 + f_1( x_{i1} ) + f_2( x_{i2} ) \ldots + f_p( x_{ip} ) + \epsilon_i = \beta_0 + \sum_{j=1}^p{f_j( x_{ij} )}  + \epsilon_i
$$

## Our Working Dataset: mpg

```{r, results = 'asis',echo=FALSE}
knitr::kable(head(mpg[,-c(1:2)]), format = "latex")
```

```{r}
mod2 <- gam(hw.mpg ~ s(weight) + s(length) , data = mpg, method = "REML")
```

```{r, fig.width=10, echo=FALSE}
plot(mod2, page = 1, shade = TRUE, residuals = TRUE, cex=1.2,shift = coef(mod2)[1], rug=FALSE)
```

### Linear Terms I

```{r,eval=FALSE}
mod3 <- gam(hw.mpg ~ s(weight) + length , data = mpg, method = "REML")
```

```{r, fig.width=10, echo=FALSE}
#mod3 <- gam(hw.mpg ~ s(weight) + s(length,k=0,fx=TRUE) , data = mpg, method = "REML")
mod3 <- gam(hw.mpg ~ s(weight) + s(length,sp=1000) , data = mpg, method = "REML")
plot(mod3, page = 1, shade = TRUE, residuals = TRUE, cex=1.2,shift = coef(mod3)[1], rug=FALSE)
```
### Linear Terms II

```{r,eval=TRUE}
mod4 <- gam(hw.mpg ~ s(weight) + s(length,sp=100) , data = mpg, method = "REML")
```

```{r, fig.width=10, echo=FALSE}
#mod3 <- gam(hw.mpg ~ s(weight) + s(length,k=0,fx=TRUE) , data = mpg, method = "REML")

plot(mod4, page = 1, shade = TRUE, residuals = TRUE, cex=1.2,shift = coef(mod4)[1], rug=FALSE)
```
### Categorical Terms I

```{r,eval=TRUE}
mod5 <- gam(hw.mpg ~ s(weight) + fuel , data = mpg, method = "REML")
```

```{r, fig.width=10, echo=FALSE}
#mod3 <- gam(hw.mpg ~ s(weight) + s(length,k=0,fx=TRUE) , data = mpg, method = "REML")

plot(mod5, page = 1, shade = TRUE, residuals = TRUE,pch=1, cex=1,shift = coef(mod5)[1], rug=FALSE, all.terms = TRUE)
```

\LARGE

In this model the nonlinear effects of weight is the same for both fuel types. (You coulod view this as the analogy of the fixed slope, different intercepts in linear models!)



### Categorical Terms II

We can also specify different smooths for different levels of teh categorical variable! ("a factor smooth interaction")

```{r,eval=TRUE}
mod6 <- gam(hw.mpg ~ s(weight, by= fuel) + fuel, data = mpg, method = "REML")
```

```{r, fig.width=10, echo=FALSE}
#mod3 <- gam(hw.mpg ~ s(weight) + s(length,k=0,fx=TRUE) , data = mpg, method = "REML")

plot(mod6, page = 1, shade = TRUE, residuals = TRUE, cex=2,shift = coef(mod6)[1], rug=FALSE)
```


## [GAM Exercises part II](exercises/GAM-Exercises-II.html)

# Interpreting GAM outputs

```{r}
mod_city4 <- gam(hw.mpg ~ s(weight) +  s(price) + s(rpm) + s(comp.ratio) + s(width) 
                 + fuel , data = mpg, method = "REML")
summary(mod_city4)

```

#### Effective Degrees of Freedom

```{r, fig.width=10, echo=FALSE}
par(mfrow=c(1,2))
plot(mod_city4, select = 1, shade = TRUE, residuals = TRUE, cex=1.2,shift = coef(mod_city4)[1], rug=FALSE)
plot(mod_city4, select = 4, shade = TRUE, residuals = TRUE, cex=1.2,shift = coef(mod_city4)[1], rug=FALSE)
```


#### Significance of Smooth Terms

Overall significance of smooth. p-values are approximate.

One way to interpret significance: you cannot draw a horizontal line through the 95% confidence interval.

```{r, fig.width=10, echo=FALSE}
par(mfrow=c(1,2))
plot(mod_city4, select = 1, shade = TRUE, residuals = TRUE, cex=1.2,shift = coef(mod_city4)[1], rug=FALSE)
abline(h= coef(mod_city4)[1],col=2)
plot(mod_city4, select = 2, shade = TRUE, residuals = TRUE, cex=1.2,shift = coef(mod_city4)[1])
abline(h= coef(mod_city4)[1],col=2, rug=FALSE)
```

#### Significance and Effective Degress of Freedom

Two separate concepts, one does not imply the other !

```{r, results = 'asis',echo=FALSE}
knitr::kable(summary(mod_city4)$s.table, format = "latex")
```

```{r, fig.width=10, echo=FALSE}
par(mfrow=c(1,3))
plot(mod_city4, select = 2, shade = TRUE, residuals = TRUE, cex=1.2,shift = coef(mod_city4)[1], rug=FALSE)
abline(h= coef(mod_city4)[1],col=2)
plot(mod_city4, select = 3, shade = TRUE, residuals = TRUE, cex=1.2,shift = coef(mod_city4)[1], rug=FALSE)
abline(h= coef(mod_city4)[1],col=2)
plot(mod_city4, select = 4, shade = TRUE, residuals = TRUE, cex=1.2,shift = coef(mod_city4)[1], rug=FALSE)
abline(h= coef(mod_city4)[1],col=2)
```



# Visualizing GAMs

```{r}
gam_mod <- gam(hw.mpg ~ s(weight) +  s(rpm) + s(price) + s(comp.ratio) , data = mpg, method = "REML")
```


```{r, fig.width=10}
plot(gam_mod, pages=1, rug=FALSE)
```


```{r, fig.width=10}
par(mfrow=c(2,2))
plot(gam_mod, select = 1, shade = TRUE, residuals = TRUE)
plot(gam_mod, select = 1, shade = TRUE, residuals = TRUE, shade.col ="lightblue")
plot(gam_mod, select = 4, shade = TRUE, residuals = TRUE, shade.col ="lightblue", rug=FALSE,pch=1)
plot(gam_mod, select = 4, shade = TRUE, residuals = TRUE, shade.col ="lightgreen", rug=FALSE,shift = coef(gam_mod)[1])
```


```{r}
plot(mod_city4, select = 1, shade = TRUE, residuals = TRUE, cex=1.2,shift = coef(mod_city4)[1], rug=FALSE)
```

# Model checking, Concurvity

\includepdf[pages={26-31,34-}]{../datacamp/chapter2.pdf}



# The gam library

In order to fit more general sorts of GAMs, using smoothing splines
or other components that cannot be expressed in terms of basis functions
and then fit using least squares regression, we will need to use the gam
library in R.

## Smoothing/number of knots specified by user

A "simpler" version of mgcv. (no Bayesian smoothing)
Note the slightly different syntax though (e.g., *df=4*, instead of *k=4*) !!

The `s()` function, which is part of the gam library, is used to indicate that
we would like to use a smoothing spline. We specify that the function of
year should have 4 degrees of freedom, and that the function of *age* will
have 5 degrees of freedom. Since *education* is qualitative, we leave it as is,
and it is converted into four dummy variables. We use the `gam()` function in
order to fit a GAM using these components. 

```{r, fig.width=10,out.width='.95\\linewidth'}
gam1 <- lm(wage ~ ns(year, 4) + ns(age, 5) + education, data = Wage)

detach("package:mgcv", unload=TRUE)#avoid name space collision!
library(gam)
gam.m3 <- gam(wage ~ s(year, 4) + s(age, 5) + education, data = Wage)
par(mfrow = c(1, 3));plot(gam.m3, se = TRUE, col = "blue")

```

Conveniently, even though gam1 is not of class gam but rather of class `lm`, we can still use `plot.gam()` on it. 

```{r, fig.width=10,out.width='.95\\linewidth'}
par(mfrow = c(1, 3));plot.Gam(gam1, se = TRUE, col = "red")
```

In these plots, the function of year looks rather linear. We can perform a
series of ANOVA tests in order to determine which of these three models is
best: a GAM that excludes year (M1), a GAM that uses a linear function
of year (M2), or a GAM that uses a spline function of year (M3).

```{r}
gam.m1 <- gam(wage ~ s(age, 5) + education, data = Wage)
gam.m2 <- gam(wage~ year + s(age, 5) + education, data = Wage)
anova(gam.m1, gam.m2, gam.m3, test = "F")
```

We find that there is compelling evidence that a GAM with a linear function
of year is better than a GAM that does not include year at all
(p-value=0.00014). However, there is no evidence that a non-linear function
of year is needed (p-value=0.349). In other words, based on the results
of this ANOVA, M2 is preferred.
The summary() function produces a summary of the gam fit.

```{r}
summary(gam.m3)
```

The p-values for *year* and *age* correspond to a null hypothesis of a linear
relationship versus the alternative of a non-linear relationship. The large
p-value for *year* reinforces our conclusion from the ANOVA test that a linear
function is adequate for this term. However, there is very clear evidence
that a non-linear term is required for *age*.
We can make predictions from *gam* objects, just like from lm objects,
using the predict() method for the class *gam*. Here we make predictions on
the training set.

```{r}
preds <- predict(gam.m2, newdata = Wage)
```

### Logistic regression

In order to fit a logistic regression GAM, we once again use the `I()` function
in constructing the binary response variable, and set `family=binomial`.


```{r, fig.width=10,out.width='.95\\linewidth'}
gam.lr.s <- gam(I(wage > 250) ~ year + s(age, df = 5) + education,
                family = binomial, data = Wage, 
                subset = (education != "1. < HS Grad"))
par(mfrow = c(1, 3)); plot(gam.lr.s, se = TRUE, col = "green")
```

# Exercises

\benum

\item \textbf{Exercises, Coding}

This question relates to the "College" data set.

\benum

\item Split the data into a training set and a test set. Using out-of-state tuition as the response and the other variables as the predictors, perform forward stepwise selection on the training set in order to identify a satisfactory model that uses just a subset of the predictors.

```{r, echo = TRUE, fig.width=10,out.width='.95\\linewidth'}
library(leaps)
set.seed(1)
attach(College)
train <- sample(length(Outstate), length(Outstate) / 2)
test <- -train
College.train <- College[train, ]
College.test <- College[test, ]
fit <- regsubsets(Outstate ~ ., data = College.train, nvmax = 17, method = "forward")
fit.summary <- summary(fit)
par(mfrow = c(1, 3))
plot(fit.summary$cp, xlab = "Number of variables", ylab = "Cp", type = "l")
min.cp <- min(fit.summary$cp)
std.cp <- sd(fit.summary$cp)
abline(h = min.cp + 0.2 * std.cp, col = "red", lty = 2)
abline(h = min.cp - 0.2 * std.cp, col = "red", lty = 2)
plot(fit.summary$bic, xlab = "Number of variables", ylab = "BIC", type='l')
min.bic <- min(fit.summary$bic)
std.bic <- sd(fit.summary$bic)
abline(h = min.bic + 0.2 * std.bic, col = "red", lty = 2)
abline(h = min.bic - 0.2 * std.bic, col = "red", lty = 2)
plot(fit.summary$adjr2, xlab = "Number of variables", ylab = "Adjusted R2", type = "l", ylim = c(0.4, 0.84))
max.adjr2 <- max(fit.summary$adjr2)
std.adjr2 <- sd(fit.summary$adjr2)
abline(h = max.adjr2 + 0.2 * std.adjr2, col = "red", lty = 2)
abline(h = max.adjr2 - 0.2 * std.adjr2, col = "red", lty = 2)
```

*Cp, BIC and adjr2 show that size 6 is the minimum size for the subset for which the scores are within 0.2 standard devitations of optimum.*

```{r, echo = TRUE}
fit <- regsubsets(Outstate ~ ., data = College, method = "forward")
coeffs <- coef(fit, id = 6)
names(coeffs)
```

\item Fit a GAM on the training data, using out-of-state tuition as the response and the features selected in the previous step as the predictors. Plot the results, and explain your findings.

\item Evaluate the model obtained on the test set, and explain the results obtained.

\item For which variables, if any, is there evidence of a non-linear relationship with the response ?

\eenum


\eenum

## [GAM Exercises part III](exercises/GAM-Exercises-CrimeData.html)

\newpage
# Appendix I

## Locations of knots

For penalized regression spline, the exact locations are not important, as long as:

* k is adequately big;
* the spread of knots has good, reasonable coverage.

By default:

* natural cubic regression spline bs = 'cr' places knots by quantile;
* B-splines family (bs = 'bs', bs = 'ps', bs = 'ad') place knots evenly.

Compare the following:

```{r, fig.width=10,out.width='.95\\linewidth'}
library(mgcv)

## toy data
set.seed(0); x <- sort(rnorm(400, 0, pi))  ## note, my x are not uniformly sampled
set.seed(1); e <- rnorm(400, 0, 0.4)
y0 <- sin(x) + 0.2 * x + cos(abs(x))
y <- y0 + e

## fitting natural cubic spline
cr_fit <- gam(y ~ s(x, bs = 'cr', k = 20))
cr_knots <- cr_fit$smooth[[1]]$xp  ## extract knots locations

## fitting B-spline
bs_fit <- gam(y ~ s(x, bs = 'bs', k = 20))
bs_knots <- bs_fit$smooth[[1]]$knots  ## extract knots locations

## summary plot
par(mfrow = c(1,2))
plot(x, y, col= "grey", main = "natural cubic spline");
lines(x, cr_fit$linear.predictors, col = 2, lwd = 2)
abline(v = cr_knots, lty = 2)
plot(x, y, col= "grey", main = "B-spline");
lines(x, bs_fit$linear.predictors, col = 2, lwd = 2)
abline(v = bs_knots, lty = 2)
```


## Setting your own knots locations:

You can also provide your customized knots locations via the knots argument of gam() (yes, knots are not fed to s(), but to gam()). For example, you can do evenly spaced knots for cr:

```{r}
xlim <- range(x)  ## get range of x
myfit <- gam(y ~ s(x, bs = 'cr', k =20),
         knots = list(x = seq(xlim[1], xlim[2], length = 20)))
```


Now you can see that:

```{r}
my_knots <- myfit$smooth[[1]]$xp
plot(x, y, col= "grey", main = "my knots");
lines(x, myfit$linear.predictors, col = 2, lwd = 2)
abline(v = my_knots, lty = 2)
```



